---
title: "factor_mutate"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{factor_mutate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

An [`R`](https://www.r-project.org) [`dplyr`](https://CRAN.R-project.org/package=dplyr) user must, to some extent, be able to know if a `dplyr` expression is well-formed.  The `dplyr` implementation makes such decisions, so to program `dplyr` one has to be able to anticipate a least a reasonable subset of these determinations.

Take the following `dplyr` statement as an example:

```{r eval = FALSE}
dplyr::mutate(x = 1,
              v1 = x,
              x = x + 1,
              v2 = x,
              x = x + 1,
              v3 = x)
```

Would the above be, in the context of a larger `dplyr` pipeline, be a valid statement?

My personal opinion is: [it is not](http://www.win-vector.com/blog/2017/09/my-advice-on-dplyrmutate/).

The issues include: the statement is (unsafely) using values in the same 
block they are created, and also re-using variable names across terms. I don't recall
these issues being discussed in [`dplyr` `mutate()` documentation](http://dplyr.tidyverse.org/reference/mutate.html), and indeed the
statement *appears* to work when used with the in-memory (or on `data.frame`)
version of `dplyr`.

```{r}
library("dplyr")
packageVersion("dplyr")

d <- data.frame(z = 1)

d %>%
  mutate(x = 1,
         v1 = x,
         x = x + 1,
         v2 = x,
         x = x + 1,
         v3 = x) %>%
  knitr::kable()
```

However "in memory" is not the only place `dplyr` is promoted and used. 
`dplyr` is also often used to talk to `SQL` back ends (such as `Spark` and 
databases).  When we try the same `mutate()` in a database context we, currently,
get a different (and wrong) result.  This establishes the example 
`mutate()` statement is at best unsafe.

```{r}
packageVersion("dbplyr")
packageVersion("DBI")
packageVersion("RSQlite")

db <- DBI::dbConnect(RSQLite::SQLite(), 
                     ":memory:")
d2 <- dplyr::copy_to(db, d, 'd2')

d2 %>%
  mutate(x = 1,
         v1 = x,
         x = x + 1,
         v2 = x,
         x = x + 1,
         v3 = x) %>%
  collect() %>%
  knitr::kable()
```

The above is an incorrect result, with no warning and no error signaled.  
Essentially this is a silent result corruption.

Looking further we can see the issue is likely in the `SQL` generation path:

```{r, comment=""}
d2 %>%
  mutate(x = 1,
         v1 = x,
         x = x + 1,
         v2 = x,
         x = x + 1,
         v3 = x) %>%
  show_query()
```

We have a function in development, `seplyr::factor_mutate()` which inspects the terms one is
considering placing in a `dplyr::mutate()` (or even a `dplyr::sumarize()` or `dplyr::transmute()`) and:

  * Issues a warning if the statement is unsafe.
  * Factors the statement into multiple ordered safe statements.

```{r}
# # seplyr::factor_mutate() currently 
# # requires the dev-version of seplyr
# # version 0.5.2 or better.
# devtools::install_github("WinVector/seplyr")
stmts <- seplyr::factor_mutate(
  x = 1,
  v1 = x,
  x = x + 1,
  v2 = x,
  x = x + 1,
  v3 = x
)
```

```{r, comment=""}
cat(stmts)
```

`seplyr::factor_mutate()` issues a warning.  It also builds statement source code
that is safe in the sense that:

 * Term to term dependencies now all cross `mutate()` stages.
 * Terms in a single `mutate()` stage can be evaluated in any order (without changing results).
 
The statement is partitioned into a minimal (subject to some order constraints)
number of new statements (though it is not obvious from this example). In particular
statements that can be left alone are left alone (and no warning is issued).

And in fact this code returns the same result on the database as
we did on the in-memory example.  


```{r}
eval(parse(text = paste("d2 %>%", stmts))) %>%
  collect() %>%
  knitr::kable()
```

We don't suggest calling `eval()` in production code.  Instead we suggest using
`seplyr::factor_mutate()` to inspect code as you are writing, and to incorporate
its advice in your best practices.

The important distinction is: the in-memory
code seems to have (possibly undocumented) strict sequential semantics (terms are guaranteed to be evaluated in order
in a single `mutate()`), and database implementations do not have such a property. Thus
we need to organize code to be completely unambiguous even with terms re-ordered 
with a *single* `mutate()` stage.  This is a service that the `seplyr::factor_mutate()`
method attempts to heuristically supply, and something *central* to the design of [`rquery`](https://winvector.github.io/rquery/)
(a `SQL`-only data processing system).

```{r cleanup}
DBI::dbDisconnect(db)
```

