---
title: "Mutate Partitioner"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mutate Partitioner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[`seplyr::partition_mutate_qt()`](https://winvector.github.io/seplyr/reference/partition_mutate_qt.html) is
a new service supplied by development version the [`R`](https://www.r-project.org) package [`seplyr`](https://winvector.github.io/seplyr/) (version `0.1.7` or higher).

`seplyr::partition_mutate_qt()` can partition a sequence of assignments so that no statement is using any value created in the same partition element or group.
The partitions are in a format accepted by [`seplyr::mutate_se()`](https://winvector.github.io/seplyr/reference/mutate_se.html) for execution.

For such a partition the evaluation result does not depend on the order of execution of the statements in each group (as they are all independent of each other's left-hand-sides).  A no-dependency small number of 
groups partition is *very* helpful when executing expressions on `SQL` based 
data interfaces (such as `Apache Spark`).

The method used to partition expressions is
to scan the remaining expressions in order taking any that: have all their values available from earlier groups and
do not use a value formed in the current group. 

This partitioning method can lead to far fewer
groups than the straightforward method of breaking up the sequence of expressions
at each new-value use.

Here is a non-trivial example (notice we use `:=` for assignment, that is a requirement of `seplyr::mutate_se()`):

```{r ex1}
library("seplyr")

plan <- partition_mutate_qt(
  rand_a := rand(),
   choice_a := rand_a>=0.5, # first use of a new value 1
    a_1 := ifelse(choice_a, # first use of a new value 2
                  'treatment', 
                  'contol'),
    a_2 := ifelse(choice_a, 
                  'control', 
                  'treatment'),
  rand_b := rand(),
   choice_b := rand_b>=0.5, # first use of a new value 3
    b_1 := ifelse(choice_b, # first use of a new value 4
                  'treatment', 
                  'contol'),
    b_2 := ifelse(choice_b, 
                  'control', 
                  'treatment'),
  rand_c := rand(),
   choice_c := rand_c>=0.5, # first use of a new value 5
    c_1 := ifelse(choice_c, # first use of a new value 6
                  'treatment', 
                  'contol'),
    c_2 := ifelse(choice_c, 
                  'control', 
                  'treatment'),
  rand_d := rand(),
   choice_d := rand_d>=0.5, # first use of a new value 7
    d_1 := ifelse(choice_d, # first use of a new value 8
                  'treatment', 
                  'contol'),
    d_2 := ifelse(choice_d, 
                  'control', 
                  'treatment'),
  rand_e := rand(),
   choice_e := rand_e>=0.5, # first use of a new value 9
    e_1 := ifelse(choice_e, # first use of a new value 10
                  'treatment', 
                  'contol'),
    e_2 := ifelse(choice_e, 
                  'control', 
                  'treatment')
  )

print(plan)
```

Notice `seplyr::partition_mutate_qt()` split the work into 3 groups.
The straightforward method (with no statement re-ordering) of splitting into
non-dependent groups would have to split the mutate at each first use of a 
new value: yielding 10 splits or 11 mutate stages.  For why a low number
of execution stages is 
important please see [here](http://winvector.github.io/FluidData/partition_mutate.html).

To execute the statements on a data-item "`d`" (either an in-memory `data.frame`
or a remote database or `Sparklyr` handle) we would do something like the following:

```{r exec, eval=FALSE}
res <- d
for(stepi in plan) {
  res <- mutate_se(res, 
                   stepi, 
                   splitTerms = FALSE)
}
```

A fully worked version of this example can be found [here](http://winvector.github.io/FluidData/partition_mutate_ex2.html).

