---
title: "Named Map Builder"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Named Map Builder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[`wrapr`](https://github.com/WinVector/wrapr) introduces an operator called "named map builder" that is written as "`:=`".  Named map builder is a *very* simple bit of code that performs a very simple task: it adds names to vectors or lists (making them work more like maps).

Here are some examples:

```{r ex1}
library("wrapr")

'a' := 5

c('a' := 5, 'b' := 6)

c('a', 'b') := c(5, 6)
```

The left-side argument of the `:=` operator is called "the names", and the right-side argument is called "the values".  The `:=` operators returns the values with the names set to names.

A key use of the named map builder is the following:

```{r key1}
key = 'keycode'
key := 'value'
```

Notice the value inside the variable `key` was used as the array name, this differs from 
what is easily done with `R`'s native `c(key = 'value')` style notation.

A great use of the `:=` operator is using it to conveniently build arguments lists for functions such as `seplyr::mutate_se()`.  This works for simple explicit code such as the following.

```{r ex2}
library("seplyr")

datasets::iris %.>%
  summarize_se(., "Mean_Sepal_Length" := "mean(Sepal.Length)")
```

Slightly more complicated code such as:

```{r ex3}
datasets::iris %.>%
  group_by_se(., "Species") %.>%
  summarize_se(., c("Mean_Sepal_Length" := "mean(Sepal.Length)",
                    "Mean_Sepal_Width" := "mean(Sepal.Width)"))
```

Or even parametric code such as:

```{r ex4}
resultColumn <- "summary_value"
datasets::iris %.>%
  group_by_se(., "Species") %.>%
  summarize_se(., resultColumn := "mean(Sepal.Length)")
```



For more details please see: [`help(`:=`, package = 'wrapr')`](https://winvector.github.io/wrapr/reference/named_map_builder.html) and 
[`help("%.>%", package="wrapr")`](https://winvector.github.io/wrapr/reference/dot_arrow.html).

