---
title: "rename_se"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rename_se}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[`seplyr`](https://github.com/WinVector/seplyr) is an attempt to work with the [`dplyr` `0.7.*`]( https://CRAN.R-project.org/package=dplyr) [`R`](https://www.r-project.org) package through standard evaluation interfaces with a minimum of change and minimum of cognitive friction.  Beyond championing standard evaluation it attempts to introduce as few of its own opinions as possible.

A case in point is `dplyr::rename()`/`seplyr::rename_se()`.  `dplyr::rename()` is used as follows.

```{r rename1}
suppressPackageStartupMessages(library("dplyr"))

datasets::mtcars %>%
  rename(cylinders = cyl, gears = gear) %>%
  head()
```

Notice `dplyr::rename()` renamings are written as assignments making them very similar to `dplyr::mutate()`.  This is the reverse from how `R` usually handles list or mapping structures.
In named vectors or named lists keys are written on the left and values are written on the right as follows.

```{r map1}
mp <- c("cyl" = "cylinders", "gear" = "gears")
print(mp)
```

Because `seplyr` is intended to be a set of adapters for `dplyr` we simply adopt `dplyr::rename()`'s convention.  This allows the user to mechanically translate their experience and expectations from scripting over `dplyr::rename()` to directly use `seplyr::rename_se()` as follows.

```{r rename2}
library("seplyr")

datasets::mtcars %.>%
  rename_se(., c("cylinders" := "cyl", "gears" := "gear")) %.>%
  head(.)
```

We hope this makes it easy to translate one-off analyses into re-usable scripts by incrementally replacing known variable names with parametric versions.  The `:=` operator is just a convenience function for building up maps, we could also have written `rename_se(c("cylinders" = "cyl", "gears" = "gear"))` or passed in a named vector built up elsewhere.

`rename_se` interpets all left-hand names as new column names and all right-hand names as old column names.  This allows `rename_se` to be used to swap columns:

```{r swap}
data.frame(a = 1, b = 2) %.>%
  rename_se(., c('a', 'b') := c('b', 'a'))
```

Please see [`help("%.>%", package="wrapr")`](https://winvector.github.io/wrapr/reference/grapes-.-greater-than-grapes.html) for details on "dot pipe."



