---
title: "seplyr"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The [`R`](https://www.r-project.org) package [`seplyr`](https://github.com/WinVector/seplyr) supplies improved *s*tandard *e*valuation interfaces for some common data data plying tasks.

To install this packing in `R` please either install from [CRAN](https://CRAN.R-project.org/package=seplyr) with:

```{r, eval=FALSE}
   install.packages('seplyr')
```

or from [`GitHub`](https://github.com/WinVector/seplyr):
 
```{r, eval=FALSE}
   devtools::install_github('WinVector/seplyr')
```

In `dplyr` if you know the names of columns when you are writing code you can write code such as the following.

```{r ex1a}
suppressPackageStartupMessages(library("dplyr"))
packageVersion("dplyr")

datasets::mtcars %>% 
  arrange(cyl, desc(gear)) %>% 
  head()
```

In `dplyr` `0.7.*` if the names of the columns are coming from a variable set elsewhere you would to need to use a tool to substitute those names in. One such tool is `rlang`/`tidyeval` (though we strongly prefer [`seplyr`](https://github.com/WinVector/seplyr/blob/master/README.md).

```{r ex1b}
# Assume this is set elsewhere,
# supplied by a user, function argument, or control file.
orderTerms <- c('cyl', 'desc(gear)')
```

If you don't want to try and digest entire theory of quasi-quoting and splicing (the `!!!` operator) then you can use `seplyr` which conveniently and legibly wraps the operations as follows:

```{r ex1c}
library("seplyr")

datasets::mtcars %.>% 
  arrange_se(., orderTerms) %>% 
  head(.)
```

The idea is: the above code looks very much like simple `dplyr` code used running an analysis, and yet is very easy to [parameterize](https://win-vector.com/2016/12/03/parametric-variable-names-and-dplyr/) and re-use in a script or package.

******

`seplyr::arrange_se()` performs the wrapping for you without you having to work through the details of `rlang`.  If you are interested in the details `seplyr` itself is a good tutorial.  For example you can examine `seplyr`'s implementation to see the necessary notations (using a command such as `print(arrange_se)`). And, of course, we try to supply some usable help entries, such as: `help(arrange_se)`.  Some more discussion of the ideas can be found [here](https://win-vector.com/2017/07/15/dplyr-0-7-made-simpler/).

The current set of SE adapters includes (all commands of the form `NAME_se()` being adapters for a `dplyr::NAME()` method):

 * `add_count_se()`
 * `add_tally_se()`
 * `arrange_se()`
 * `count_se()`
 * `distinct_se()`
 * `filter_se()`
 * `group_by_se()`
 * `group_indices_se()`
 * `mutate_se()`
 * `rename_se()`
 * `select_se()`
 * `summarize_se()`
 * `tally_se()`
 * `transmute_se()`

Only two of the above are completely redundant. `seplyr::group_by_se()` essentially works as `dplyr::group_by_at()` and `seplyr::select_se()` essentially works as `dplyr::select_at()`. The others either have different semantics or currently (as of `dplyr` `0.7.1`) no matching `dplyr::*_at()` method.  Roughly all `seplyr` is trying to do is give a uniform first-class standard interface to all of the primary deprecated underscore suffixed verbs (such as `dplyr::arrange_`).

We also have a few methods that work around a few of the minor inconvenience of working with variable names as strings:

* `deselect()`
* `rename_mp()`

Here is a example using `seplyr::summarize_se()`.

```{r atexse}
datasets::iris %.>%
  group_by_se(., "Species") %.>%
  summarize_se(., c("Mean.Sepal.Length" := "mean(Sepal.Length)", 
                    "Mean.Sepal.Width" := "mean(Sepal.Width)"))
```

In addition to the series of adapters we also supply a number of useful new verbs including:

 * `group_summarize()` Binds grouping, arrangement, and summarization together for clear documentation of intent.
 * `add_group_summaries()` Adds per-group summaries to data.
 * `add_group_indices()` Adds a column of per-group ids to data.
 * `add_group_sub_indices()` Adds a column of in-group rank ids to data.
 * `add_rank_indices()` Adds rank indices to data.
 * [`partition_mutate_se()`](https://winvector.github.io/seplyr/reference/partition_mutate_se.html): [vignette](https://winvector.github.io/seplyr/articles/MutatePartitioner.html), and [article](https://winvector.github.io/FluidData/partition_mutate.html).
 * [`if_else_device()`](https://winvector.github.io/seplyr/reference/if_else_device.html): [article](https://winvector.github.io/FluidData/partition_mutate_ex2.html).


`seplyr` is designed to be a thin package that passes work to `dplyr`.  If you want a package that works around `dplyr` implementation differences on different data sources I suggest trying our own [`replyr`]( https://CRAN.R-project.org/package=replyr) package. Another alternative is using [`wrapr::let()`](https://github.com/WinVector/wrapr/blob/master/README.md).

`seplyr` methods are short and have examples in their help, so always try both help and printing the method (for example: `help(select_se)` and `print(select_se)`). Printing methods can show you how to use `dplyr` directly with `rlang`/`tidyeval` methodology (allowing you to skip `seplyr`).

Some inspiration comes from [Sebastian Kranz's `s_dplyr`](https://gist.github.com/skranz/9681509). 
Please see [`help("%.>%", package="wrapr")`](https://winvector.github.io/wrapr/reference/dot_arrow.html) for details on "dot pipe."


