---
title: "String Algebra"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{String Algebra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[`seplyr`](https://winvector.github.io/seplyr/index.html) supplies the function 
`seplyr::expand_expr()` which implements what we call "the string algebra"
or string expression interpolation.
The function takes an expression of mixed terms, including: variables
referring to names, quoted strings, and general expression terms.  It then "de-quotes"
all of the variables referring to quoted strings and  "dereferences" variables 
thought to be referring to names.  The entire expression is then returned as a single string.

This is easiest to see with an example:

```{r stra1}
library("seplyr")

ratio <- 2
compCol1 <- "Sepal.Width"
expr <- expand_expr("Sepal.Length" >= ratio * compCol1)
print(expr)
```

`expand_expr` works by capturing the user supplied expression unevaluated, performing some transformations, and returning the entire expression as a single quoted string (essentially
returning new source code).

Notice in the above one layer of quoting was removed from `"Sepal.Length"` and the name
referred to by "`compCol1`" was substituted into the expression. "`ratio`" was left
alone as it was not referring to a string (and hence can not be a name; unbound or free variables
are also left alone).  So we  see that the substitution performed does depend on what values
are present in the environment.

If you want to be stricter in your specification, you could add quotes around any symbol you do not want de-referenced.
For example:

```{r expr2}
expand_expr("Sepal.Length" >= "ratio" * compCol1)
```

After the substitution the returned quoted expression is exactly in the form `seplyr` expects.
For example:

```{r mu1}
resCol1 <- "Sepal_Long"

datasets::iris %.>%
  mutate_se(., 
            resCol1 := expr) %.>%
  head(.)
```

Details on [`%.>%`](https://winvector.github.io/wrapr/reference/grapes-.-greater-than-grapes.html) (dot pipe) and [`:=`](https://winvector.github.io/wrapr/reference/named_map_builder.html) (named map builder) can be found [here](https://winvector.github.io/wrapr/articles/dot_pipe.html) and [here](https://winvector.github.io/seplyr/articles/named_map_builder.html) respectively.  The idea is:
`seplyr::mutate_se(., "Sepal_Long" := "Sepal.Length >= ratio * Sepal.Width")` should be equilant to `dplyr::mutate(., Sepal_Long = Sepal.Length >= ratio * Sepal.Width)`.


`seplyr` also provides an number of `seplyr::*_nse()` convenience forms wrapping all of these steps into one operation.  For example:

```{r mu2}
datasets::iris %.>%
  mutate_nse(., 
             resCol1 := "Sepal.Length" >= ratio * compCol1) %.>%
  head(.)
```

To use string literals you merely need one extra layer of quoting:

```{r mu3}
"is_setosa" := expand_expr(Species == "'setosa'")
datasets::iris %.>%
  transmute_nse(., 
             "is_setosa" := Species == "'setosa'") %.>%
  summary(.)
```

The purpose of all of the above is to mix names that are known while we are 
writing the code (these are quoted) with names that may not be known until later
(i.e., column names supplied as parameters).  This allows the easy creation of 
useful generic functions such as:

```{r f1}
countMatches <- function(data, columnName, targetValue) {
  # extra quotes to say we are interested in value, not de-reference
  targetSym <- paste0('"', targetValue, '"') 
  data %.>%
    transmute_nse(., "match" := columnName == targetSym) %.>%
    group_by_se(., "match") %.>%
    summarize_se(., "count" := "n()")
}

countMatches(datasets::iris, "Species", "setosa")
```

The purpose of the `seplyr` string system is to pull off quotes and de-reference indirect variables.  So, you need to remember to add enough extra quotation marks to prevent this where you do not want it.



