---
title: "Using seplyr to Program Over dplyr"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using seplyr to Program Over dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

[`seplyr`](https://github.com/WinVector/seplyr) is an [`R`](https://www.r-project.org) package that makes it easy to program over [`dplyr` `0.7.*` +]( https://CRAN.R-project.org/package=dplyr) without needing to directly use `rlang` notation.


## `seplyr`

`seplyr` is a `dplyr` adapter layer that prefers "slightly clunkier" standard interfaces (or [referentially transparent](https://en.wikipedia.org/wiki/Referential_transparency) interfaces), which are actually very powerful and can be used to some advantage.

The above description and comparisons can come off as needlessly broad and painfully abstract.  Things are much clearer if we move away from theory and return to our example.

Let's translate the above example into a re-usable function in small (easy) stages.  First translate the interactive script from `dplyr` notation into `seplyr` notation.  This step is a pure [re-factoring](https://en.wikipedia.org/wiki/Code_refactoring), we are changing the code without changing its observable external behavior.

The translation is mechanical in that it is mostly using `seplyr` documentation as a lookup table.  What you have to do is:

 * Change `dplyr` verbs to their matching `seplyr` "`*_se()`" adapters.
 * Add quote marks around names and expressions.
 * Convert sequences of expressions (such as in the `summarize()`) to explicit vectors by adding the "`c()`" notation.
 * Replace "`=`" in expressions with "`:=`".
 
Our converted code looks like the following.


```{r exc}
library("dplyr")
```

```{r}
library("seplyr")
```

```{r}
starwars %>%
  group_by_se("homeworld") %>%
  summarize_se(c("mean_height" := "mean(height, na.rm = TRUE)",
                 "mean_mass" := "mean(mass, na.rm = TRUE)",
                 "count" := "n()"))
```

This code works the same as the original `dplyr` code.  Also the translation could be performed by following the small set of explicit re-coding rules that we gave above.

Obviously at this point all we have done is: worked to make the code a bit less pleasant looking. We have yet to see any benefit from this conversion (though we can turn this on its head and say all the original `dplyr` notation is saving us is from having to write a few quote marks).

The benefit is: this new code can *very easily* be parameterized and wrapped in a re-usable function.  In fact it is now simpler to do than to describe.


```{r exp3}
grouped_mean <- function(data, 
                         grouping_variables, 
                         value_variables,
                         count_name = "count") {
  result_names <- paste0("mean_", 
                         value_variables)
  expressions <- paste0("mean(", 
                        value_variables, 
                        ", na.rm = TRUE)")
  calculation <- result_names := expressions
  data %>%
    group_by_se(grouping_variables) %>%
    summarize_se(c(calculation,
                   count_name := "n()")) %>%
    ungroup()
}

starwars %>% 
  grouped_mean(grouping_variables = c("eye_color", "skin_color"),
               value_variables = c("mass", "birth_year"))
```

We have translated our original interactive or ad-hoc calculation into a parameterized reusable function in two easy steps:

  * Translate into \code{seplyr} notation.
  * Replace values with variables.
  
To be sure: there are some clunky details of using \code{paste0()} to build up the expressions, but the conversion process is very regular and easy. In `seplyr` parametric programming is intentionally easy (just replace values with variables).  



## Conclusion

The `seplyr` methodology is simple, easy to teach, and powerful.  

There are alternatives that differ in philosophy.

  * `rlang`/tidyeval is often used to build new non-standard interfaces on top of existing non-standard interfaces.  This is needed because non-standard interfaces do not naturally compose.  The tidy/non-tidy analogy is: it works around a mess by introducing a new mess.
  * `wrapr::let()` converts non-standard interfaces into standard [referentially transparent](https://en.wikipedia.org/wiki/Referential_transparency) or value-oriented interfaces.  It tries to help clean up messes.
  * `seplyr` is a demonstration of a possible variation of `dplyr` where more of the interfaces expect values.  It tries to cut down on the initial mess, which in turn cuts down on the need for tools and training in dealing with messes.

The `seplyr` package contains a number of worked examples both in `help()` and [`vignette(package='seplyr')`](https://winvector.github.io/seplyr/articles/seplyr.html) documentation. 



